---
title: "SCENIC analysis"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are going to analyse the results from the pySCENIC command-line software. SCENIC is a tool for single-cell regulatory network inference and clustering.

## Setup

Define chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/15-scenic-analysis.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(scater)
library(scran)
```

Load CRAN packages:

```{r}
library(ggthemes)
library(pheatmap)
library(viridis)
```

Read experiment object:

```{r}
sce <- readRDS("output/09-cell-annotation.rds")
```

## Activity

Read activity matrix:

```{r}
auc <- read.csv("data/pyscenic-activity.csv", row.names = "Regulon", stringsAsFactors = FALSE)

auc <- as.matrix(auc)
```

Create tableau colour palette:

```{r}
TableauPalette <- tableau_color_pal("Tableau 10")
```

Create cluster colour palette:

```{r}
sce$cluster <- factor(sce$cluster)

EmbryoClusterColours <- levels(sce$cluster)

EmbryoClusterColours <- TableauPalette(nlevels(sce$cluster))

names(EmbryoClusterColours) <- levels(sce$cluster)
```

Create heatmap annotations and palette:

```{r}
ann <- list(
  annotation_col = data.frame(Cluster = sce$cluster, row.names = colnames(sce)),
  annotation_colors = list(Cluster = EmbryoClusterColours)
)
```

Plot matrix of AUC values for all regulons:

```{r fig.width = 8, fig.height = 6}
pheatmap(
  mat = auc,
  color = viridis(100),
  annotation_col = ann$annotation_col,
  annotation_colors = ann$annotation_colors,
  border_color = NA,
  show_rownames = FALSE,
  show_colnames = FALSE,
)
```

Identify outlier or potentially "interesting" regulons:

```{r}
out <- isOutlier(rowVars(auc), nmads = 1, type = "higher", log = TRUE)
```

Plot matrix of AUC values for outlier regulons:

```{r fig.width = 8, fig.height = 6}
pheatmap(
  mat = auc[out, ],
  color = viridis(100),
  annotation_col = ann$annotation_col,
  annotation_colors = ann$annotation_colors,
  border_color = NA,
  show_rownames = TRUE,
  show_colnames = FALSE,
)
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
