---
title: "Cell heterogeneity"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

The cells in this sample are quite homogenous, so they do not form distinct clusters. To identify subtle differences between the cells, we can use a pre-defined set of interesting genes to examine the biological heterogeneity.

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/19-cell-heterogeneity.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(BiocSingular)
library(scater)
library(scran)
```

Load CRAN packages:

```{r}
library(ggforce)
library(msigdbr)
library(pheatmap)
```

Read experiment data:

```{r}
sce <- readRDS("output/09-cell-annotation.rds")
```

## Lineage pathways {.tabset}

Retrieve all gene sets and their member genes:

```{r}
all <- msigdbr(species = "Mus musculus")
```

Select relevant ontology terms:

```{r}
key <- c(
  "GO_ECTODERM_DEVELOPMENT",
  "GO_EPITHELIAL_TO_MESENCHYMAL_TRANSITION",
  "GO_PRIMITIVE_STREAK_FORMATION",
  "GO_PARAXIAL_MESODERM_DEVELOPMENT",
  "GO_PARAXIAL_MESODERM_FORMATION",
  "GO_PARAXIAL_MESODERM_MORPHOGENESIS",
  "GO_DETERMINATION_OF_LEFT_RIGHT_ASYMMETRY_IN_LATERAL_MESODERM",
  "GO_LATERAL_MESODERM_DEVELOPMENT",
  "GO_GASTRULATION",
  "GO_ENDOTHELIUM_DEVELOPMENT"
)
```

Subset gene sets by ontology terms:

```{r}
all <- subset(all, gs_name %in% key)
```

Plot heatmap of gene expression values:

```{r, fig.width = 8, fig.height = 15, results = "asis"}
for (k in key) {
  
  dat <- subset(all, gs_name == k)

  cat("### ", k, "\n")
  
  plotHeatmap(sce, dat$gene_symbol, center = TRUE, symmetric = TRUE, colour_columns_by = "cluster", fontsize = 7)
  
  cat("\n\n")
  
}
```

## {.tabset}

Perform PCA on expression data:

```{r}
for (k in key) {
  
  set.seed(802231141)
  
  lab <- paste("PCA", k, sep = "_")
  
  row <- subset(all, gs_name == k)
  
  dim <- calculatePCA(sce, subset_row = row$gene_symbol, BSPARAM = ExactParam())
  
  fit <- getClusteredPCs(dim, min.rank = 1, max.rank = ncol(dim))
  
  num <- metadata(fit)$chosen
  
  len <- seq_len(num)
  
  dim[, len]
  
  reducedDim(sce, lab) <- dim[, len]

}
```

Perform TSNE on PCA matrix:

```{r}
for (k in key) {
  
  set.seed(545413716)
  
  id1 <- paste("TSNE", k, sep = "_")
  
  id2 <- paste("PCA", k, sep = "_")

  reducedDim(sce, id1) <- calculateTSNE(sce, dimred = id2)

}
```

Perform UMAP on PCA matrix:

```{r}
for (k in key) {
  
  set.seed(498830989)
  
  id1 <- paste("UMAP", k, sep = "_")
  
  id2 <- paste("PCA", k, sep = "_")

  reducedDim(sce, id1) <- calculateUMAP(sce, dimred = id2)

}
```

### PCA {.tabset}

```{r, fig.width = 10, fig.height = 10, results = "asis"}
for (k in key) {
  
  var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")
  
  dim <- paste("PCA", k, sep = "_")
  
  plt <- lapply(var, function(x) plotReducedDim(sce, dimred = dim, colour_by = x) + coord_fixed(1) + theme_no_axes())

  cat("#### ", k, "\n")
  
  print(patchwork::wrap_plots(plt, ncol = 2))
  
  cat("\n\n")
  
}
```

### TSNE {.tabset}

```{r, fig.width = 10, fig.height = 10, results = "asis"}
for (k in key) {
  
  var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")
  
  dim <- paste("TSNE", k, sep = "_")
  
  plt <- lapply(var, function(x) plotReducedDim(sce, dimred = dim, colour_by = x) + theme_no_axes() + theme(aspect.ratio = 1))

  cat("#### ", k, "\n")
  
  print(patchwork::wrap_plots(plt, ncol = 2))
  
  cat("\n\n")
  
}
```

### UMAP {.tabset}

```{r, fig.width = 10, fig.height = 10, results = "asis"}
for (k in key) {
  
  var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")
  
  dim <- paste("UMAP", k, sep = "_")
  
  plt <- lapply(var, function(x) plotReducedDim(sce, dimred = dim, colour_by = x) + theme_no_axes() + theme(aspect.ratio = 1))

  cat("#### ", k, "\n")
  
  print(patchwork::wrap_plots(plt, ncol = 2))
  
  cat("\n\n")
  
}
```

## Lineage markers {.tabset}

Read lineage markers:

```{r}
ids <- read.csv("data/lineage-markers.csv")
```

Plot heatmap of gene expression values:

```{r}
plotHeatmap(sce, ids$gene_id, center = TRUE, symmetric = TRUE, colour_columns_by = "cluster")
```

Perform PCA on logcounts matrix of lineage markers:

```{r}
set.seed(802231141)

sce <- runPCA(sce, subset_row = ids$gene_id, BSPARAM = ExactParam())
```

Find the number of subpopulations in the data using clustering:

```{r}
dim <- reducedDim(sce, "PCA")

fit <- getClusteredPCs(dim, min.rank = 1, max.rank = ncol(dim))

num <- metadata(fit)$chosen
```

Choose number of PCs to retain:

```{r}
len <- seq_len(num)

dim <- reducedDim(sce, "PCA")

reducedDim(sce, "PCA") <- dim[, len]
```

Perform TSNE on PCA matrix:

```{r}
set.seed(545413716)

sce <- runTSNE(sce, dimred = "PCA")
```

Perform UMAP on PCA matrix:

```{r}
set.seed(498830989)

sce <- runUMAP(sce, dimred = "PCA")
```

### PCA

```{r fig.width = 8, fig.height = 8}
var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")

plt <- lapply(var, function(x) plotReducedDim(sce, dimred = "PCA", colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### TSNE

```{r fig.width = 8, fig.height = 8}
var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")

plt <- lapply(var, function(x) plotReducedDim(sce, dimred = "TSNE", colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### UMAP

```{r fig.width = 8, fig.height = 8}
var <- c("cluster", "phase", "celltype_aucell", "celltype_singler", "Etv2", "Kdr", "Sox2", "T")

plt <- lapply(var, function(x) plotReducedDim(sce, dimred = "UMAP", colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
