---
title: "Atlas trajectory"
author: "James Ashmore"
date: "2020-10-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/12-atlas-trajectory.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(MouseGastrulationData)
library(scater)
library(scran)
library(slingshot)
library(tradeSeq)
```

Load CRAN packages:

```{r}
library(bigmemory)
library(ggforce)
library(ggthemes)
library(pheatmap)
library(RColorBrewer)
```

Source convienence functions:

```{r}
source("code/reorder_within.R")
```

Read experiment object:

```{r}
sce <- readRDS("output/11-atlas-neighbors.rds")
```

## Diffusion

Remove cells not known to be present in the data:

```{r}
nan <- c("NMP", "PGC", "Surface ectoderm")

sce <- sce[, !sce$celltype %in% nan]
```

Produce a diffusion map for the cells:

```{r}
set.seed(385008972)

sce <- runDiffusionMap(sce, dimred = "corrected")
```

Define annotation labels for visualisation:

```{r}
source.labels <- c("reference" = "Mouse Embryo Atlas", "test" = "Mouse LPMP LHF")
```

Plot DiffusionMap of cells coloured by source data:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(DiffusionMap.1, DiffusionMap.2, colour = source)) + 
  geom_point() + 
  scale_colour_tableau(name = "Source", labels = source.labels, palette = "Tableau 10") + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() + theme(aspect.ratio = 1)
```

Plot DiffusionMap of cells coloured by embryonic stage:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(DiffusionMap.1, DiffusionMap.2, colour = stage)) + 
  geom_point() + 
  scale_colour_tableau(name = "Stage", palette = "Tableau 10", na.value = "#000000") + 
  labs(x = "DC1", y = "DC2") + 
  facet_wrap(~ source, labeller = labeller(source = source.labels)) + 
  theme_bw() + theme(aspect.ratio = 1)
```

Plot DiffusionMap of cells coloured by celltype annotation:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(DiffusionMap.1, DiffusionMap.2, colour = celltype)) + 
  geom_point() + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  labs(x = "DC1", y = "DC2") + 
  facet_wrap(~ source, labeller = labeller(source = source.labels)) + 
  theme_bw() + theme(aspect.ratio = 1)
```

## Entropy

Function to compute the entropy of the expression profile within each cell:

```{r}
whichNonZero <- function(x) {
  
  x <- as(x, "SparseArraySeed")
  
  z <- nzindex(x)
  
  list(i = z[, 1], j = z[, 2], x = nzdata(x))

}

calculateEntropy <- function(x) {
  
  n <- whichNonZero(counts(x))
  
  p <- n$x / colSums(counts(x))[n$j]
  
  e <- - p * log(p)
  
  e <- by(e, INDICES = n$j, FUN = sum)
  
  e <- as.vector(e)
  
}
```

Calculate the entropy of the expression profile within each cell:

```{r}
sce$entropy <- calculateEntropy(sce)
```

Plot DiffusionMap of cells from reference and test data coloured by entropy:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(DiffusionMap.1, DiffusionMap.2, colour = entropy)) + 
  geom_point() + 
  scale_colour_viridis_c() + 
  labs(x = "DC1", y = "DC2") + 
  facet_wrap(~ source, labeller = labeller(source = source.labels)) + 
  theme_bw() + theme(aspect.ratio = 1)
```

Plot DiffusionMap of cells from reference data coloured by entropy:

```{r fig.width = 8, fig.height = 6}
ggcells(sce[, sce$source == "reference"], aes(DiffusionMap.1, DiffusionMap.2, colour = entropy)) + 
  geom_point() + 
  scale_colour_viridis_c() + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() + theme(aspect.ratio = 1)
```

Plot DiffusionMap of cells from test data coloured by entropy:

```{r fig.width = 8, fig.height = 6}
ggcells(sce[, sce$source == "test"], aes(DiffusionMap.1, DiffusionMap.2, colour = entropy)) + 
  geom_point() + 
  scale_colour_viridis_c() + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() + theme(aspect.ratio = 1)
```

Plot entropy of cells from reference and test data:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(x = reorder_within(x = celltype, by = -entropy, within = stage, fun = median), y = entropy, colour = celltype)) + 
  geom_sina(seed = 42) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  facet_wrap(~ stage, scales = "free_x") + 
  labs(y = "Entropy", colour = "Celltype") + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

Plot entropy of cells from reference data:

```{r fig.width = 8, fig.height = 6}
ggcells(sce[, sce$source == "reference"], aes(x = reorder_within(x = celltype, by = -entropy, within = stage, fun = median), y = entropy, colour = celltype)) + 
  geom_sina(seed = 42) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  facet_wrap(~ stage, scales = "free_x") + 
  labs(y = "Entropy", colour = "Celltype") + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

Plot entropy of cells from test data:

```{r fig.width = 8, fig.height = 6}
ggcells(sce[, sce$source == "test"], aes(x = reorder_within(x = celltype, by = -entropy, within = stage, fun = median), y = entropy, colour = celltype)) + 
  geom_sina(seed = 42) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  facet_wrap(~ stage, scales = "free_x") + 
  labs(y = "Entropy", colour = "Celltype") + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

## Trajectory

Function to calculate distance with MNN coordinates:

```{r}
dist_clusters_diag <- function(X, w1, w2){
    if(length(w1) != nrow(X) | length(w2) != nrow(X)){
        stop("Reduced dimensional matrix and weights vector contain different
             numbers of points.")
    }
    mu1 <- colWeightedMeans(X, w = w1)
    mu2 <- colWeightedMeans(X, w = w2)
    diff <- mu1 - mu2
    if(sum(w1>0)==1){
        s1 <-  diag(ncol(X))
    }else{
        s1 <- diag(diag(cov.wt(X, wt = w1)$cov))
    }
    if(sum(w2>0)==1){
        s2 <-  diag(ncol(X))
    }else{
        s2 <- diag(diag(cov.wt(X, wt = w2)$cov))
    }
    return(as.numeric(t(diff) %*% solve(s1 + s2) %*% diff))
}
```

Perform lineage inference with Slingshot:

```{r}
set.seed(378891116)

sce <- slingshot(sce, reducedDim = "corrected", dist.fun = dist_clusters_diag)
```

Reverse pseudotime to fit with known developmental trajectory:

```{r}
sce$slingPseudotime_1 <- max(abs(sce$slingPseudotime_1)) + -sce$slingPseudotime_1
```

Construct path data from trajectory curves:

```{r}
sc1 <- embedCurves(sce, "DiffusionMap")

sc1 <- slingCurves(sc1)[[1]]

sc1 <- data.frame(sc1$s[sc1$ord, ])
```

Plot DiffusionMap of cells coloured by pseudotime:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(DiffusionMap.1, DiffusionMap.2)) + 
  geom_point(aes(colour = slingPseudotime_1)) + 
  geom_path(data = sc1, aes(DC1, DC2), size = 0.8) + 
  scale_colour_viridis_c() + 
  labs(x = "DC1", y = "DC2", colour = "Pseudotime") + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

Plot pseudotime of cells coloured by celltype and facteted by data source:

```{r fig.width = 8, fig.height = 6}
ggcells(sce, aes(slingPseudotime_1, reorder(celltype, slingPseudotime_1, mean), colour = celltype)) + 
  geom_jitter(show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours) + 
  facet_wrap(~ source, labeller = labeller(source = source.labels)) + 
  labs(x = "Pseudotime", y = "Celltype", colour = "Celltype") + 
  theme_bw() + theme(aspect.ratio = 1)
```

## Differential

Fit the NB-GAM model with batch effect design matrix on highly variable genes:

```{r}
set.seed(430396281)

mod <- model.matrix(~ 0 + source, colData(sce))

colnames(mod) <- c("reference", "test")

idx <- which(rowData(sce)$gene_id %in% metadata(sce)$getTopHVGs)

fit <- fitGAM(counts = counts(sce), sds = SlingshotDataSet(sce), U = mod, genes = idx)
```

Test whether average gene expression is associated with pseudotime:

```{r}
res <- associationTest(fit)

res$FDR <- p.adjust(res$pvalue, method = "fdr")

res$symbol <- rowData(sce)$gene_name[match(rownames(res), rowData(sce)$gene_id)]

res <- subset(res, FDR < 0.05)

res <- res[order(res$pvalue), ]
```

Create table of associationTest results:

```{r}
knitr::knit_print(DT::datatable(res))
```

Function to plot expression of genes associated with pseudotime:

```{r}
plotPseudotime <- function(object, features) {
  
  colnames(object) <- paste("cell", seq_len(ncol(object)), sep = "-")
  
  object <- object[features, order(object$slingPseudotime_1)]
  
  x <- assay(object, "reconstructed")
  
  rownames(x) <- rowData(object)$gene_name

  M <- rowMeans(x, na.rm = TRUE)
  
  DF <- ncol(x) - 1L
  
  IsNA <- is.na(x)
  
  if (any(IsNA)) {
    
    mode(IsNA) <- "integer"
    
    DF <-  DF - rowSums(IsNA)
    
    DF[DF == 0L] <- 1L
  
  }

  x <- x - M
  
  V <- rowSums(x^2L, na.rm = TRUE) / DF
	
  x <- x / sqrt(V + 0.01)
  
  pheatmap.cluster_rows <- hclust(dist(x, method = "euclidean"), method = "complete")

  pheatmap.color <- colorRampPalette(rev(brewer.pal(n = 5, name = "RdBu")))(100)
  
  l <- max(abs(x))
  
  pheatmap.breaks <- seq(-l, l, length.out = 101)
  
  pheatmap.annotation_col <- data.frame(
    Celltype = object$celltype, 
    Stage = object$stage,
    Source = object$source,
    row.names = colnames(object))
  
  pheatmap.annotation_colors = list(
    Celltype = EmbryoCelltypeColours[names(EmbryoCelltypeColours) %in% object$celltype],
    Stage = setNames(tableau_color_pal("Tableau 10")(5), sort(unique(sce$stage))),
    Source = setNames(tableau_color_pal("Classic Green-Orange 6")(2), sort(unique(sce$source)))
  )

  pheatmap(
    mat = x,
    color = pheatmap.color,
    breaks = pheatmap.breaks,
    cluster_rows = pheatmap.cluster_rows,
    cluster_cols = FALSE,
    annotation_col = pheatmap.annotation_col,
    annotation_colors = pheatmap.annotation_colors,
    show_colnames = FALSE
  )

}
```

Plot expression of genes associated with pseudotime:

```{r fig.width = 12, fig.height = 8}
plotPseudotime(sce, head(rownames(res), n = 50))
```

## Summary

### Output

Save integrated experiment object to disk:

```{r}
saveRDS(sce, file = "output/12-atlas-trajectory.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
