---
title: "Interactive data exploration"
author: "James Ashmore"
date: "2020-10-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Load CRAN packages:

```{r}
library(rsconnect)
```

Include Bioconductor package repositories:

```{r}
options(repos = BiocManager::repositories())
```

Automatically answer yes to any user prompt:

```{r}
options(needs.promptUser = FALSE)
```

Create shiny directory for interactive data exploration:

```{r}
dir.create("shiny", showWarnings = FALSE, mode = "0755")
```

## Applications

### Processed data

Create iSEE directory for processed data:

```{r}
dir.create("shiny/v1", showWarnings = FALSE, mode = "0755")
```

Copy processed data to iSEE directory:

```{r}
file.copy("output/09-cell-annotation.rds", "shiny/v1/data.rds", overwrite = TRUE)
```

Write application code to file within iSEE directory:

```{cat, engine.opts = list(file = "shiny/v1/app.R"), class.source = "r"}
library(SingleCellExperiment)

library(BiocSingular)

library(scran)

sce <- readRDS("data.rds")

library(iSEE)

library(shiny)

packageVersion("iSEE")

celltype_colours = function(n) {

  col <- c("Allantois" = "#532C8A",
           "Anterior Primitive Streak" = "#c19f70",
           "Blood progenitors 1" = "#f9decf",
           "Blood progenitors 2" = "#c9a997",
           "Cardiomyocytes" = "#B51D8D",
           "Caudal epiblast" = "#9e6762",
           "Caudal Mesoderm" = "#3F84AA",
           "Caudal neurectoderm" = "#354E23",
           "Def. endoderm" = "#F397C0",
           "Endothelium" = "#ff891c",
           "Epiblast" = "#635547",
           "Erythroid1" = "#C72228",
           "Erythroid2" = "#f79083",
           "Erythroid3" = "#EF4E22",
           "ExE ectoderm" = "#989898",
           "ExE endoderm" = "#7F6874",
           "ExE mesoderm" = "#8870ad",
           "Forebrain/Midbrain/Hindbrain" = "#647a4f",
           "Gut" = "#EF5A9D",
           "Haematoendothelial progenitors" = "#FBBE92",
           "Intermediate mesoderm" = "#139992",
           "Mesenchyme" = "#cc7818",
           "Mixed mesoderm" = "#DFCDE4",
           "Nascent mesoderm" = "#C594BF",
           "Neural crest" = "#C3C388",
           "NMP" = "#8EC792",
           "Notochord" = "#0F4A9C",
           "Paraxial mesoderm" = "#8DB5CE",
           "Parietal endoderm" = "#1A1A1A",
           "PGC" = "#FACB12",
           "Pharyngeal mesoderm" = "#C9EBFB",
           "Primitive Streak" = "#DABE99",
           "Rostral neurectoderm" = "#65A83E",
           "Somitic mesoderm" = "#005579",
           "Spinal cord" = "#CDE088",
           "Surface ectoderm" = "#f7f79e",
           "Visceral endoderm" = "#F6BFCB"
  )

  return(col)

}

ecm <- ExperimentColorMap(colData = list(celltype = celltype_colours))

options(iSEE.maxlevels = 37)

iSEE(sce, appTitle = "LPMP_EHF_B", colormap = ecm)
```

Deploy the iSEE application to the shiny server:

```{r}
deployApp(appDir = "shiny/v1", appName = "LPMP_EHF_B", appTitle = "LPMP_EHF_B", launch.browser = FALSE, forceUpdate = TRUE)
```

Allocate more memory to the iSEE application:

```{r}
configureApp(appName = "LPMP_EHF_B", appDir = "shiny/v1", size = "xlarge")
```

### Integrated data

Create iSEE directory for integrated data:

```{r}
dir.create("shiny/v2", showWarnings = FALSE, mode = "0755")
```

Copy integrated data to iSEE directory:

```{r}
file.copy("output/14-atlas-markers.rds", "shiny/v2/data.rds", overwrite = TRUE)
```

Write application code to file within iSEE directory:

```{cat, engine.opts = list(file = "shiny/v2/app.R"), class.source = "r"}
library(SingleCellExperiment)

library(BiocSingular)

library(scran)

sce <- readRDS("data.rds")

library(iSEE)

library(shiny)

packageVersion("iSEE")

celltype_colours = function(n) {

  col <- c("Allantois" = "#532C8A",
           "Anterior Primitive Streak" = "#c19f70",
           "Blood progenitors 1" = "#f9decf",
           "Blood progenitors 2" = "#c9a997",
           "Cardiomyocytes" = "#B51D8D",
           "Caudal epiblast" = "#9e6762",
           "Caudal Mesoderm" = "#3F84AA",
           "Caudal neurectoderm" = "#354E23",
           "Def. endoderm" = "#F397C0",
           "Endothelium" = "#ff891c",
           "Epiblast" = "#635547",
           "Erythroid1" = "#C72228",
           "Erythroid2" = "#f79083",
           "Erythroid3" = "#EF4E22",
           "ExE ectoderm" = "#989898",
           "ExE endoderm" = "#7F6874",
           "ExE mesoderm" = "#8870ad",
           "Forebrain/Midbrain/Hindbrain" = "#647a4f",
           "Gut" = "#EF5A9D",
           "Haematoendothelial progenitors" = "#FBBE92",
           "Intermediate mesoderm" = "#139992",
           "Mesenchyme" = "#cc7818",
           "Mixed mesoderm" = "#DFCDE4",
           "Nascent mesoderm" = "#C594BF",
           "Neural crest" = "#C3C388",
           "NMP" = "#8EC792",
           "Notochord" = "#0F4A9C",
           "Paraxial mesoderm" = "#8DB5CE",
           "Parietal endoderm" = "#1A1A1A",
           "PGC" = "#FACB12",
           "Pharyngeal mesoderm" = "#C9EBFB",
           "Primitive Streak" = "#DABE99",
           "Rostral neurectoderm" = "#65A83E",
           "Somitic mesoderm" = "#005579",
           "Spinal cord" = "#CDE088",
           "Surface ectoderm" = "#f7f79e",
           "Visceral endoderm" = "#F6BFCB"
  )

  return(col)

}

ecm <- ExperimentColorMap(colData = list(celltype = celltype_colours))

options(iSEE.maxlevels = 37)

iSEE(sce, appTitle = "LPMP_EHF_B_ATLAS", colormap = ecm)
```

Deploy the iSEE application to the shiny server:

```{r}
deployApp(appDir = "shiny/v2", appName = "LPMP_EHF_B_ATLAS", appTitle = "LPMP_EHF_B_ATLAS", launch.browser = FALSE, forceUpdate = TRUE)
```

Allocate more memory to the iSEE application:

```{r}
configureApp(appName = "LPMP_EHF_B_ATLAS", appDir = "shiny/v2", size = "xlarge")
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
